---
title: "STAT-670-Assignment 1"
author: 'Prashanth Thirukkurungudi Sekar (UID: 2000376633)'
date: "January 16, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Worked with Pravin Sundar and Barathwaaj Parthasarathy \



## Loading the libraries

```{r}
library(ggplot2)
library(readr)
library(stringr)
```
## Reading the data

```{r}

nba2017_18 <- read_csv("nba2017-18.csv")
pacers2017_18 = read_csv("pacers2017-18.csv")
```

## Cleaning the name variable

```{r}
pacers2017_18 = data.frame(pacers2017_18)
pacers2017_18$Player = as.character(pacers2017_18$Player)
pacers2017_18$Player = str_split_fixed(pacers2017_18$Player, "\\\\", 2)[,1]
```

## Creating height variable with inches

```{r}
pacers2017_18$Height = as.numeric(substring(pacers2017_18$Ht,1,1)) * 12 + as.numeric(substring(pacers2017_18$Ht,3,nchar(pacers2017_18$Ht)))
```

In order to improve scatter plot provided, a horizontal histogram visualization is used with the names in the y-axis and the heights (in inches) in the x-axis. \




## Improvisation of the plot

```{r}
o = order(pacers2017_18$Height)
pacers2017_18$Player = factor(pacers2017_18$Player, levels = pacers2017_18$Player[order(pacers2017_18$Pos,pacers2017_18$Height)])
gg = ggplot(pacers2017_18, aes(x = Player, y = Height, fill = Pos)) + geom_bar(stat="identity") + coord_flip()
gg + xlab("Height (inches)") + ggtitle("Heights of the 2017-18 Pacers")

```

The names are ordered by position first and then according to the heights within each position. Using this visualization, it is easier to observe that "Centers" are the tallest in the team and the "Guards" being the shortest in the team.