---
title: "Project Code"
author: "Prashanth"
date: "January 31, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the libraries and data: 

```{r}
library(haven)
library(dplyr)
library(tidyr)
library(viridis)
LM_WPID_web_2 <- read_dta("LM_WPID_web_2.dta")
ventile_2011_for_release_LCU <- read_dta("ventile_2011_for_release_LCU.dta")
wyd_88_05_for_release <- read_dta("wyd_88_05_for_release.dta")
```

## Question 1

Subsetting only for US 

```{r}
WPID_US <- LM_WPID_web_2 %>% filter(country == "United States")
head(WPID_US)
```

```{r}
WPID_US$group <- factor(WPID_US$group)
head(WPID_US)
#View(WPID_US)
```

```{r}
library(ggplot2)
cb_palette = c("#CD6155", "#AF7AC5", "#7FB3D5", "#76D7C4", "#F7DC6F", "#EDBB99", "#B3B6B7", "#1C2833", "#D55E00", "#CC79A7")
ggplot(WPID_US,aes(x = bin_year, y = RRinc, color = group)) + geom_line() + xlab("Year") + ylab("Per capita income (in USD)") + ggtitle("Per capita income between deciles in US") + geom_point() + scale_x_continuous("Year", labels = as.character(WPID_US$bin_year), breaks = WPID_US$bin_year) + scale_color_manual(values = cb_palette)
```

```{r}
library(ggplot2)
library(forecast)
library(dplyr)

WPID_US_yoy <- WPID_US %>% dplyr::select(country, bin_year, group, RRinc)
WPID_US_yoy <- WPID_US_yoy %>% spread(bin_year, RRinc)
WPID_US_yoy$`88_93` <- (WPID_US_yoy$`1993`- WPID_US_yoy$`1988`)*100/WPID_US_yoy$`1988`
WPID_US_yoy$`93_98` <- (WPID_US_yoy$`1998`- WPID_US_yoy$`1993`)*100/WPID_US_yoy$`1993`
WPID_US_yoy$`98_03` <- (WPID_US_yoy$`2003`- WPID_US_yoy$`1998`)*100/WPID_US_yoy$`1998`
WPID_US_yoy$`03_08` <- (WPID_US_yoy$`2008`- WPID_US_yoy$`2003`)*100/WPID_US_yoy$`2003`

WPID_US_yoy_final <- WPID_US_yoy %>% dplyr::select(country,group,`88_93`,`93_98`,`98_03`,`03_08`)
WPID_US_yoy_final <- WPID_US_yoy_final %>% gather(Year_to_Year,YOY_Change,c(`88_93`,`93_98`,`98_03`,`03_08`))

WPID_US_yoy_final <- WPID_US_yoy_final[order(WPID_US_yoy_final$group),]
WPID_US_yoy_final$group <- factor(WPID_US_yoy_final$group)

WPID_US_yoy_final$order1 = ifelse(WPID_US_yoy_final$Year_to_Year=="88_93",1,ifelse(WPID_US_yoy_final$Year_to_Year=="93_98",2
                                                                                  ,ifelse(WPID_US_yoy_final$Year_to_Year=="98_03",3,4)))

o = order(WPID_US_yoy_final$order1)
WPID_US_yoy_final$Year_to_Year = factor(WPID_US_yoy_final$Year_to_Year, levels = unique(WPID_US_yoy_final$Year_to_Year[o]))

ggplot(WPID_US_yoy_final,aes(x = Year_to_Year, y = YOY_Change, color = WPID_US_yoy_final$group ,group = WPID_US_yoy_final$group)) + geom_line() + geom_point()  + xlab("Year to Year") + ylab("Per capita income change (in USD)") + ggtitle("Per capita income change with decile in US")+ scale_color_manual(values = cb_palette)
```


```{r}
ggplot(WPID_US_yoy_final,aes(x = Year_to_Year, y = YOY_Change, color = WPID_US_yoy_final$group ,group = WPID_US_yoy_final$group)) + geom_line() + geom_point()  + xlab("Year to Year") + ylab("Percentage Change") + ggtitle("Per capita income change for each decile in US")+ ylim(c(-5,30))+ theme(legend.position="none") +
  facet_wrap(~WPID_US_yoy_final$group,ncol=5) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + geom_hline(aes(yintercept=0)) + scale_color_manual(values = cb_palette)
```

Getting numerical results :

```{r}
tt = WPID_US_yoy
tt = tt[,c(1:7)]
tt$`1988_1993` = round(((tt$`1993` - tt$`1988`)*1)/(tt$`1988`),4)
tt$`1993_1998` = round(((tt$`1998` - tt$`1993`)*1)/(tt$`1993`),4)
tt$`1998_2003` = round(((tt$`2003` - tt$`1998`)*1)/(tt$`1998`),4)
tt$`2003_2008` = round(((tt$`2008` - tt$`2003`)*1)/(tt$`2003`),4)
tt$`2008_1988` = round(((tt$`2008` - tt$`1988`)*1)/(tt$`2008`),4)
tt1 = tt[,c(1,2,8:12)]
```

Getting 2nd numerical results :

```{r warning=FALSE, message=FALSE}
tt = WPID_US_yoy
tt = tt[,c(1:7)]
library(reshape2)
tt = reshape(tt, direction = "long", varying = list(names(tt)[3:7]), v.names = "Value", 
        idvar  = c("country","group"), timevar = "Year", times = c(1988,1993,1998,2003,2008))

tt = tt[,c(1,3,2,4)]

library(data.table)
tt = data.table(tt)
tt2 = tt[ , diff := Value - shift(Value), by = Year]    

tt2 = tt2[,c(2,3,5)]
library(reshape)
tt2 <- cast(tt2, Year~group,mean)
first = tt$Value[tt$group==1]
tt2$`1` = first
tt2$last_decile_minus_first_decile = tt2$`10`-tt2$`1`
```

## Question 2

Subsetting the required 5 countries

```{r}
d0 = ventile_2011_for_release_LCU[ventile_2011_for_release_LCU$contcod %in% c('USA','GBR','DEU','ZAF','AUS'),]
```

Taking absolute difference from country's mean

```{r}
library(dplyr)
d1 = d0 %>% left_join(d0 %>% group_by(contcod) %>% summarise(Mean = mean(ventile_income)),by=c("contcod"))
d1$mean_diff = d1$ventile_income / d1$Mean
```

Plotting the result

```{r}
library(ggplot2)
color_map = viridis(5)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")
ggplot(d1,aes(x=ventile,y=mean_diff,group = contcod,color=contcod)) +geom_line() + geom_path(size = 1)+ ggtitle("Distribution of Income relative to Country's Mean") + labs(x="Ventiles",y = "Ratio between Income and Mean")+ scale_x_continuous("Ventiles", labels = as.character(d1$ventile), breaks = d0$ventile) + guides(color=guide_legend(title="Country Codes")) + geom_hline(aes(yintercept=1)) + scale_color_manual(values=cb_palette)# scale_color_viridis(discrete=TRUE)
```

Looking the distribution of incomes relative to the country mean across ventiles for the countries, Austrailia(AUS), Germany(DEU), United Kingdom(GBR), United States(USA) and South Africe(ZAF), we observe the following:
1. South Africa has the highest relative income amongst the countries in the last ventile.
2. Austrailia has the lowest relative income at ventile 1, meaning, among all the income generated by 5% of individuals in a country with the lowest income(ventile 1), Austrailia has the least income compared to its mean.
3. It can be seen that Germany and United Kingdom have very similar pattern, being almost parralel to each other. These two countries seem to have a very balanced distribution across the mean and the ventiles. The difference in the magnitude of income generated by the induviduals having least income (ventile 1) in the country is not drastically different from the highest earning individual(ventile 20); which can be seen in the case of AUS USA and ZAF. 
4. Austrailia and USA have a very similar pattern only varying at extremes, where Austrailia has a lower income at ventile 1 and has a higher income at ventile 20 having a more wider range of incomes around its mean.
5. It is interesting to note that the mean income for South Africa occurs at ventile 15-60 but for all other countries it occurs at ventile 13. Leading to derive a conclution that the income generated in ZAF across ventiles is comparitively higher compared to the other 4 countries. 



## Question 3

```{r}
ventile_v1 = ventile_2011_for_release_LCU
ventile_v2 = ventile_v1[ventile_v1$contcod %in% c("CHN-R","CHN-U"),] %>% group_by(ventile) %>% summarise(Sum_income=sum(ventile_income))
ventile_v2$contcod = "CHN"
ventile_v1 = ventile_v1[!(ventile_v1$contcod %in% c("CHN-R","CHN-U")), ]
ventile_v1 = ventile_v1[,c(1,5,6)]
ventile_v2 = ventile_v2[,c(3,1,2)]
names(ventile_v2) = names(ventile_v1)
ventile_v3 = rbind(ventile_v1,ventile_v2)
```


```{r}
d2 = ventile_v3 %>% left_join(ventile_v3 %>% group_by(contcod) %>% summarise(Mean = mean(ventile_income),Sum=sum(ventile_income)),by=c("contcod"))
d2$percentage = (d2$ventile_income/d2$Sum)*100
d3 = d2[d2$ventile %in% c(1,20),]
lower = d3[d3$ventile==1,]
upper = d3[d3$ventile==20,]
lower$Mean_log = log1p(lower$Mean)
upper$Mean_log = log1p(upper$Mean)
```

```{r}
ggplot(lower,aes(y=percentage,x=Mean_log)) + geom_point() + geom_smooth(method="lm",se=FALSE) + labs(y="Percentage of Income (Ventile)",x="Logged Value of mean income of the country") + ggtitle("Percentage of Income earned by Bottom 5% with Mean income of the country")
```

```{r}
ggplot(upper[upper$percentage<60,],aes(y=percentage,x=Mean_log)) + geom_point() + geom_smooth(method="lm",se=FALSE) + labs(y="Percentage of Income (Ventile)",x="Logged Value of mean income of the country") + ggtitle("Percentage of Income earned by Top 5% with Mean income of the country")
```


```{r}
library(lattice)
library(viridis)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
d2 = ventile_v3 %>% left_join(ventile_v3 %>% group_by(contcod) %>% summarise(Mean = mean(ventile_income),Sum=sum(ventile_income)),by=c("contcod"))
d2$percentage = (d2$ventile_income/d2$Sum)*100
d3 = d2[d2$ventile %in% c(1,20),]
continents = read.csv("continents.csv")
d3 = data.frame(d3)
continents$Continent.Code = as.character(continents$Continent.Code)
continents$contcod = as.character(continents$contcod)
continents = continents[complete.cases(continents),]
d4 = left_join(d3,continents,by="contcod")
d4$Continent.Code = ifelse(is.na(d4$Continent.Code),"NA",d4$Continent.Code)

lower = d4[d4$ventile==1,]
upper = d4[d4$ventile==20,]
lower$Mean_log = log1p(lower$Mean)
upper$Mean_log = log1p(upper$Mean)
```



```{r}
ggplot(lower,aes(y=percentage,x=Mean_log,color=Continent.Code)) + geom_point() + geom_smooth(method="lm",se=FALSE,aes(group=1),colour="black") + labs(y="Percentage of Income",x="Logged Value of mean income of the country") + ggtitle("Percentage of Income earned by Bottom 5% with Mean income of the country") +  scale_colour_manual(values =cb_palette)
```
```{r}
ggplot(upper[upper$percentage<60,],aes(y=percentage,x=Mean_log,color=Continent.Code)) + geom_point() + geom_smooth(method="lm",se=FALSE,aes(group=1),colour="black") + labs(y="Percentage of Income",x="Logged Value of mean income of the country") + ggtitle("Percentage of Income earned by Top 5% with Mean income of the country")+
  scale_colour_manual(values =cb_palette)
```

```{r}
lower.lm = lm((Mean_log)~(percentage), data = lower)
#library(broom)
#lower.lm.df = augment(lower.lm)
#library(arm)
#display((lower.lm.df))
```

```{r}
summary(lower.lm)
```

```{r}
upper.lm = lm((Mean_log)~(percentage), data = upper[upper$percentage<60,])

```

```{r}
summary(upper.lm)
```

Plotting World Map :

```{r}


lower1 = lower
lower1$contcod = ifelse(lower1$contcod=="CHN-R","CHN",lower1$contcod)
upper1 =upper
upper1$contcod = ifelse(upper1$contcod=="CHN-R","CHN",upper1$contcod)
```

```{r}
library(rworldmap)
malMap <- joinCountryData2Map(lower, joinCode = "ISO3",
                              nameJoinColumn = "contcod")
library(RColorBrewer)
library(viridis)
colourPalette <- viridis(100)#brewer.pal(5,'RdYlGn')
mapCountryData(malMap, nameColumnToPlot="percentage", catMethod = "numerical",colourPalette = colourPalette,
               missingCountryCol = gray(.8),mapTitle = "Percentage of Income earned by Bottom 5%")

```

```{r}

malMap <- joinCountryData2Map(upper[upper$percentage<60,], joinCode = "ISO3",
                              nameJoinColumn = "contcod")

colourPalette <- viridis(100)#brewer.pal(5,'RdYlGn')
mapCountryData(malMap, nameColumnToPlot="percentage", catMethod = "numerical",colourPalette = colourPalette,
               missingCountryCol = gray(.8),mapTitle = "Percentage of Income earned by Top 5%")

```

Looking at the two plots, we can see that there are three types of countries. The first set are Southeast Asian countries and some central African countries. These countries correspond to high percentage of income generated by the bottom 5% of the population(Pakistan,Israel,Armenia,Egypt,etc.,). We can see that these countries are nowhere near the top in percentage of income generated by top 5% of the population. \

The second set of countries are those which are among the top in percentage of income generated by top 5% of the population. These are generally South American and South African countries. The percentage of income generated by the bottom 5% of the population of these countries are among the least. This suggests that there is imbalance in incomes of the people. Rich get richer, poor get poorer. \

The third type of countries are those which are moderate in both categories. These are the countries which have balanced income distribution among the ventiles(Jordan,India,Nicaragua,etc.,).