---
title: "STAT-670-Assignment 2"
author: 'Prashanth Thirukkurungudi Sekar (UID: 2000376633)'
date: "January 23, 2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading the libraries

```{r results='hide', message=FALSE, warning=FALSE}
library(readr)
library(tidyr)
library(ggplot2)
library(lattice)
library(reshape)
library(dplyr)
```
## Reading the data

```{r message=FALSE}
anes_pilot_2016 <- read_csv("anes_pilot_2016.csv")
anes_pilot_2016 = data.frame(anes_pilot_2016)
```

Subsetting the data :

```{r message=FALSE}
required_cols = c('version','caseid','fttrump','fthrc','ftsanders','ftrubio','immig_numb')
d0 = anes_pilot_2016[,required_cols]
```

Removing outlier values for "feeling thermometer" measurements:

```{r message=FALSE}
d1 = d0[d0$fttrump<=100,]
d1 = d1[d1$fthrc<=100,]
d1 = d1[d1$ftsanders<=100,]
d1 = d1[d1$ftrubio<=100,]
```

Checking if all outliers are removed:

```{r message=FALSE}
summary(d1)
```

## Question 1

Formatting the data :

```{r message=FALSE}
d2 = melt(d1[,c(2:7)],id=c("caseid","immig_numb"))
names(d2) = c("caseid","immig_numb","candidate","value")
d2$candidate = ifelse(d2$candidate=="fttrump","Donald Trump",ifelse(d2$candidate=="fthrc","Hillary Clinton",ifelse(d2$candidate=="ftsanders","Bernie Sanders","Marco Rubio")))

```

```{r message=FALSE, warning=FALSE}

neworder <- c("Donald Trump","Hillary Clinton","Bernie Sanders","Marco Rubio")
d2 <- arrange(mutate(d2,
             candidate=factor(candidate,levels=neworder)),candidate)

ggplot(d2,aes(x=value)) + geom_density(fill="#0072B2") + ggtitle("Feeling Thermometer measurements for four Presidential Candidates\n (0-Very Cold and 100-Very Warm)") + labs(x="Feeling Thermometer Measurements",y="Density") + 
  facet_wrap(~candidate,ncol=2) + theme_bw()

```

#### Description:
It can be observed that both Hillary Clinton and Donald Trump have received a lot of "Very Cold" (Negative responses) scores and this number is higher than their respective  number of "Very Warm" (Positive responses) scores. Their distributions are very identical and most of the people were able to decide conclusively on their attitude towards both the candidates (Either positive/negative). \
Bernie Sanders had the highest number of "Very Warm" scores (positive responses) when compared with all the candidates. The distribution has 3 peaks (trimodal)- at 0 (Very Cold), 100 (Very Warm) and 50 (Moderate) and all the 3 peaks are almost similar and it can be inferred that there was an equal share of people having positive, negative and neutral opinion about his candidacy.\
While for Marco Rubio, the distribution is more of a bimodal distribution with peaks at 0 (Very Cold) and 50 (Moderate). In comparison with Clinton and Trump, Rubio has fewer "Very Cold" scores (negative responses).Amongst all candidates, he has the highest "Moderate" scores (Score of 50) and the least "Very Warm" scores (positive responses). One reason for a lot of scores in the range of 50 could be that people were not very sure about his candidacy (or) had a neutral opinion about his candidacy.


## Question 2

Formatting the data :

```{r message=FALSE}
d3 = d2 %>% group_by(immig_numb,candidate) %>% summarise(Mean = mean(value))
d3$immig_numb = as.factor(d3$immig_numb)
d3 = d3[order(d3$candidate,d3$immig_numb),]
```


```{r}
ploarity = c("#0A2229","#0174DF","#81DAF5","#A4A4A4","#F78181","#FF0000","#610B0B")
neworder <- c("Bernie Sanders","Hillary Clinton","Donald Trump","Marco Rubio")
d4 <- arrange(mutate(d3,
             candidate=factor(candidate,levels=neworder)),candidate)
ggplot(d4,aes(x=immig_numb,y=Mean,color = immig_numb)) + geom_point(size=3.5) +facet_wrap(.~candidate,ncol=2) +labs(x="Attitude towards Immigration",y="Mean feeling thermometer score") + ggtitle("Mean thermometer score for each candidate based on Immigration Attitude") + theme_bw() + 
scale_color_manual(values=ploarity,name="Immigration Attitude",labels = c("Increased a lot","Increased a moderate amount","Increased a little","Kept the same","Decreased a little","Decreased a moderate amount","Decreased a lot"))
```


#### Description:

For Hillary Clinton and Bernie Sanders, people who wanted the immigration number to be increased gave high score for "feeling thermometer" measure (positive responses) and thereby supported their candidacy. While people who wanted to decrease the immigration number were not in support of them resulting in a low mean "feeling thermometer" score. \
While for Donald Trump, people who wanted the immigration number to be decreased are those who supported him (or) rather gave high scores for "feeling thermometer" measure. While people who wanted an increase in the immigration number were not supportive of him thereby resulting in a low mean "feeling thermometer" score.\
For Marco Rubio, the people who wanted the immigration number to be increased were not supporting him (low mean "feeling thermometer" scores), while the people who wanted to decrease the immigration number had an indifferent attitude towards his candidacy resulting in mean "feeling thermometer" scores around the range of 50 (Moderate Scores).

Note : Given that "Immigration Policy" was a major talking point during the 2016 election campaign, it is plausible that the people who had a positive belief about immigration, gave high scores for candidates who were supportive of increase in immigration, while people who were against immigartion gave high scores for candidates who were not supportive of increase in immigration.



